---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# premp

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/premp)](https://CRAN.R-project.org/package=premp)
<!-- badges: end -->

**Pre**dictive **M**astery **P**rofiles (PreMP) is a method for setting achievement standards designed to be compatible with assessments scored using diagnostic classification models.
The goal of premp is to provide functionality to implement this method in practice.

## Installation

You can install the development version of premp like so:

``` r
pak::pak("r-dcm/premp")
```

## Usage

This is a basic example which shows you how to solve a common problem:

```{r example}
#library(premp)
## basic example code
```

---

## Contributions and Code of Conduct

Contributions are welcome.
To ensure a smooth process, please review the [Contributing Guide](https://premp.r-dcm.org/CONTRIBUTING.html).
Please note that the dcmstan project is released with a [Contributor Code of Conduct](https://premp.r-dcm.org/CODE_OF_CONDUCT.html).
By contributing to this project, you agree to abide by its terms.
